---
title: "Ufo Sightings in the United States"
format:
  closeread-html:
    toc: true
    self-contained: true
    cr-style:
      section-background-color: white
    code-overflow: wrap
theme:
  - custom.scss
editor: visual
---

```{r setup, include=FALSE}
library(ggdist)
library(ggplot2)
library(dplyr)
library(readr)
library(tidyverse)
library(gganimate)
library(transformr)
library(gifski)
library(maps)
library(ggmap)
library(janitor)

# this sets ggplot objects to a default theme that is relatively clean
theme_set(theme_ggdist())
```

This project analyzes over 80,000 reported UFO sightings across the United States between 1990 and 2013. By visualizing **where** these sightings occurred, we explore how cultural moments, media influence, and historical events may have shaped public perception of unidentified flying objects. Our goal is to understand the broader patterns behind these reports and the societal context in which they emerged.

```{r}
#load the CSV dataset
ufo_dataset <- read.csv("scrubbed.csv")

#clean it
ufo_by_year <- ufo_dataset %>%
  mutate(datetime = as.Date(datetime, format = "%m/%d/%Y %H:%M")) %>%
  filter(!is.na(datetime)) %>%
  mutate(year = as.integer(format(datetime, "%Y"))) %>%
  filter(year >= 1990, year <= 2020) %>%
  group_by(year) %>%
  summarise(sightings = n())

#to be used in scaling
max_sightings <- max(ufo_by_year$sightings)
```

:::::::::::::::: cr-section
Why did UFO sightings rise so dramatically from the 1990s into the early 2010s? Was it due to more activity in the skies - or more happening here on Earth? @cr-ufo-animated

::: {#cr-ufo-animated}
```{r fig.retina=1, warning=FALSE, message=FALSE}
#| fig-width: 6
#| fig-height: 5
#| fig-dpi: 100
#| 
animated_data <- ufo_by_year %>% filter(year >= 1990 & year <= 2013)

p_animated <- ggplot(animated_data, aes(x = year, y = sightings)) +
  geom_line(color = "steelblue", linewidth = 1.2) +
  scale_x_continuous(limits = c(1990, 2013), breaks = seq(1990, 2013, 5)) +
  scale_y_continuous(limits = c(0, max_sightings)) +
  labs(title = 'UFO Sightings (1990–2013)', x = 'Year', y = 'Sightings') +
  transition_reveal(year) +
  ease_aes("linear")

animate(
  p_animated,
  device = "ragg_png",
  renderer = gifski_renderer(loop = FALSE),
  width = 6, height = 5, units = "in", res = 100,
  nframes = 40, fps = 8
)
```
:::

The initial spike in 1995 sightings may reflect growing cultural obsession with aliens, sparked by [The X-Files]{style="color: green;"} (1993), the viral [Alien Autopsy]{style="color: green;"} broadcast, and rising Area 51 conspiracy chatter online. @cr-ufo-1

::: {#cr-ufo-1}
```{r fig.retina=1, warning=FALSE, message=FALSE}
#| fig-width: 6
#| fig-height: 5
#| fig-dpi: 100

ufo_by_year %>%
  filter(year <= 1995) %>%
  ggplot(aes(x = year, y = sightings, group = 1)) +
  geom_line(color = "steelblue", linewidth = 1.2) +
  scale_x_continuous(limits = c(1990, 2013), breaks = seq(1990, 2013, 5)) +
  scale_y_continuous(limits = c(0, max_sightings)) +
  labs(title = 'UFO Sightings (1990–1995)', x = 'Year', y = 'Sightings')
```
:::

After a brief dip in 1996, UFO sightings surged between 1997 and 1999, reaching new heights of public interest. @cr-ufo-2

::: {#cr-ufo-2}
```{r fig.retina=1, echo=FALSE, warning=FALSE, message=FALSE}
#| fig-width: 6
#| fig-height: 5
#| fig-dpi: 100

old_data <- ufo_by_year %>% filter(year <= 1995)
new_data <- ufo_by_year %>% filter(year >= 1995 & year <= 1999)

p2 <- ggplot() +
  geom_line(data = old_data, aes(x = year, y = sightings), color = "lightblue", linewidth = 1.2) +
  geom_line(data = new_data, aes(x = year, y = sightings), color = "steelblue", linewidth = 1.2) +
  scale_x_continuous(limits = c(1990, 2013), breaks = seq(1990, 2013, 5)) +
  scale_y_continuous(limits = c(0, max_sightings)) +
  labs(title = 'UFO Sightings (1990–1999)', x = 'Year', y = 'Sightings')

p2
```
:::

Interestingly, 1997 marked the [Phoenix Lights]{style="color: green;"} event, one of the most widely reported mass sightings in history. Around the same time, blockbuster films like [Independence Day]{style="color: green;"} (1996) and [Men in Black]{style="color: green;"} (1997) dominated pop culture, further legitimizing UFO discourse and likely fueling a wave of new reports. @cr-ufo-3

::: {#cr-ufo-3}
```{r fig.retina=1, echo=FALSE, warning=FALSE, message=FALSE}
#| fig-width: 6
#| fig-height: 5
#| fig-dpi: 100

old_data <- ufo_by_year %>% filter(year <= 1995)
new_data <- ufo_by_year %>% filter(year >= 1995 & year <= 1999)

p3 <- ggplot() +
  geom_line(data = old_data, aes(x = year, y = sightings), color = "lightblue", linewidth = 1.2) +
  geom_line(data = new_data, aes(x = year, y = sightings), color = "steelblue", linewidth = 1.2) +
  geom_vline(xintercept = 1997, linetype = "dashed", color = "darkgreen") +
  annotate("text", x = 1997, y = max_sightings * 0.95, label = "Phoenix Lights", color = "darkgreen", angle = 90, vjust = -0.5, size = 3) +
  scale_x_continuous(limits = c(1990, 2013), breaks = seq(1990, 2013, 5)) +
  scale_y_continuous(limits = c(0, max_sightings)) +
  labs(title = 'UFO Sightings (1990–1999)', x = 'Year', y = 'Sightings')

p3
```
:::

Sightings rose by over 50% over the next five years - evidence of a general upward trend heading into the early 2000s. @cr-ufo-4

::: {#cr-ufo-4}
```{r fig.retina=1, echo=FALSE, warning=FALSE, message=FALSE}
#| fig-width: 6
#| fig-height: 5
#| fig-dpi: 100

old_data <- ufo_by_year %>% filter(year <= 1999)
new_data <- ufo_by_year %>% filter(year >= 1999 & year <= 2004)

label_1999 <- ufo_by_year %>% filter(year == 1999)
label_2004 <- ufo_by_year %>% filter(year == 2004)

p4 <- ggplot() +
  geom_line(data = old_data, aes(x = year, y = sightings), color = "lightblue", linewidth = 1.2) +
  geom_line(data = new_data, aes(x = year, y = sightings), color = "steelblue", linewidth = 1.2) +
  geom_text(data = label_1999,
            aes(x = year, y = sightings, label = sightings),
            vjust = -1, color = "steelblue", size = 3) +
  geom_text(data = label_2004,
            aes(x = year, y = sightings, label = sightings),
            vjust = -1, color = "steelblue", size = 3) +
  scale_x_continuous(limits = c(1990, 2013), breaks = seq(1990, 2013, 5)) +
  scale_y_continuous(limits = c(0, max_sightings)) +
  labs(title = 'UFO Sightings (1990–2004)', x = 'Year', y = 'Sightings')

p4
```
:::

Sightings peaked in 2004, coinciding with the [USS Nimitz]{style="color: green;"} incident - Navy pilot footage of fast-moving craft that reignited public and military interest in UFOs. @cr-ufo-5

::: {#cr-ufo-5}
```{r fig.retina=1, echo=FALSE, warning=FALSE, message=FALSE}
#| fig-width: 6
#| fig-height: 5
#| fig-dpi: 100

old_data <- ufo_by_year %>% filter(year <= 1999)
new_data <- ufo_by_year %>% filter(year >= 1999 & year <= 2006)

p5 <- ggplot() +
  geom_line(data = old_data, aes(x = year, y = sightings), color = "lightblue", linewidth = 1.2) +
  geom_line(data = new_data, aes(x = year, y = sightings), color = "steelblue", linewidth = 1.2) +
  geom_vline(xintercept = 2004, linetype = "dashed", color = "darkgreen") +
  annotate("text", x = 2004, y = max_sightings * 0.88, label = "USS Nimitz Encounter", color = "darkgreen", angle = 90, vjust = -0.5, size = 3) +
  scale_x_continuous(limits = c(1990, 2013), breaks = seq(1990, 2013, 5)) +
  scale_y_continuous(limits = c(0, max_sightings)) +
  labs(title = 'UFO Sightings (1990–2006)', x = 'Year', y = 'Sightings')

p5
```
:::

Between 2006 and 2008, sightings rose sharply. Why did this sudden spike occur? @cr-ufo-6

::: {#cr-ufo-6}
```{r fig.retina=1, echo=FALSE, warning=FALSE, message=FALSE}
#| fig-width: 6
#| fig-height: 5
#| fig-dpi: 100

old_data <- ufo_by_year %>% filter(year <= 2006)
new_data <- ufo_by_year %>% filter(year >= 2006 & year <= 2008)

p6 <- ggplot() +
  geom_line(data = old_data, aes(x = year, y = sightings), color = "lightblue", linewidth = 1.2) +
  geom_line(data = new_data, aes(x = year, y = sightings), color = "steelblue", linewidth = 1.2) +
  scale_x_continuous(limits = c(1990, 2013), breaks = seq(1990, 2013, 5)) +
  scale_y_continuous(limits = c(0, max_sightings)) +
  labs(title = 'UFO Sightings (1990–2008)', x = 'Year', y = 'Sightings')

p6
```
:::

Two widely reported events likely played a role: the [2006 O'Hare Incident]{style="color: green;"}, where airport staff saw a disc-shaped object hovering over a United gate, and the [Stephenville, Texas]{style="color: green;"} sighting in early January 2008, involving dozens of credible witnesses and radar confirmation. @cr-ufo-7

::: {#cr-ufo-7}
```{r fig.retina=1, echo=FALSE, warning=FALSE, message=FALSE}
#| fig-width: 6
#| fig-height: 5
#| fig-dpi: 100

old_data <- ufo_by_year %>% filter(year <= 2006)
new_data <- ufo_by_year %>% filter(year >= 2006 & year <= 2008)

p7 <- ggplot() +
  geom_line(data = old_data, aes(x = year, y = sightings), color = "lightblue", linewidth = 1.2) +
  geom_line(data = new_data, aes(x = year, y = sightings), color = "steelblue", linewidth = 1.2) +
  geom_vline(xintercept = 2006, linetype = "dashed", color = "darkgreen", linewidth = 0.5) +
  annotate("text", x = 2006, y = max_sightings * 0.9, label = "O'Hare UFO", color = "darkgreen", angle = 90, vjust = -0.4, size = 3) +
  geom_vline(xintercept = 2008, linetype = "dashed", color = "darkgreen", linewidth = 0.5) +
  annotate("text", x = 2008, y = max_sightings * 0.9, label = "Stephenville, TX", color = "darkgreen", angle = 90, vjust = -0.4, size = 3) +
  scale_x_continuous(limits = c(1990, 2013), breaks = seq(1990, 2013, 5)) +
  scale_y_continuous(limits = c(0, max_sightings)) +
  labs(title = 'UFO Sightings (1990–2008)', x = 'Year', y = 'Sightings')

p7
```
:::

After 2008, sightings dipped slightly through 2010 - perhaps a return to baseline following the media surge around earlier high-profile events? @cr-ufo-8

::: {#cr-ufo-8}
```{r fig.retina=1, echo=FALSE, warning=FALSE, message=FALSE}
#| fig-width: 6
#| fig-height: 5
#| fig-dpi: 100

old_data <- ufo_by_year %>% filter(year <= 2008)
new_data <- ufo_by_year %>% filter(year >= 2008 & year <= 2010)

p8 <- ggplot() +
  geom_line(data = old_data, aes(x = year, y = sightings), color = "lightblue", linewidth = 1.2) +
  geom_line(data = new_data, aes(x = year, y = sightings), color = "steelblue", linewidth = 1.2) +
  scale_x_continuous(limits = c(1990, 2013), breaks = seq(1990, 2013, 5)) +
  scale_y_continuous(limits = c(0, max_sightings)) +
  labs(title = 'UFO Sightings (1990–2010)', x = 'Year', y = 'Sightings')

p8
```
:::

But from 2010 to 2012, sightings climbed again, reaching new heights and suggesting renewed public engagement with the UFO phenomenon. @cr-ufo-9

::: {#cr-ufo-9}
```{r fig.retina=1, echo=FALSE, warning=FALSE, message=FALSE}
#| fig-width: 6
#| fig-height: 5
#| fig-dpi: 100

old_data <- ufo_by_year %>% filter(year <= 2010)
new_data <- ufo_by_year %>% filter(year >= 2010 & year <= 2012)

p9 <- ggplot() +
  geom_line(data = old_data, aes(x = year, y = sightings), color = "lightblue", linewidth = 1.2) +
  geom_line(data = new_data, aes(x = year, y = sightings), color = "steelblue", linewidth = 1.2) +
  scale_x_continuous(limits = c(1990, 2013), breaks = seq(1990, 2013, 5)) +
  scale_y_continuous(limits = c(0, max_sightings)) +
  labs(title = 'UFO Sightings (1990–2012)', x = 'Year', y = 'Sightings')

p9
```
:::

The surge in sightings from 2010 to 2012 may have been driven by growing social media use, the viral spread of UFO content, and cultural phenomena like the [2012 Mayan Apocalypse]{style="color: green;"} and the launch of [Ancient Aliens]{style="color: green;"} on TV. @cr-ufo-11

::: {#cr-ufo-10}
```{r fig.retina=1, echo=FALSE, warning=FALSE, message=FALSE}
#| fig-width: 6
#| fig-height: 5
#| fig-dpi: 100

old_data <- ufo_by_year %>% filter(year <= 2010)
new_data <- ufo_by_year %>% filter(year >= 2010 & year <= 2012)

p10 <- ggplot() +
  geom_line(data = old_data, aes(x = year, y = sightings), color = "lightblue", linewidth = 1.2) +
  geom_line(data = new_data, aes(x = year, y = sightings), color = "steelblue", linewidth = 1.2) +
  geom_vline(xintercept = 2012, linetype = "dashed", color = "darkgreen") +
  annotate("text", x = 2012, y = max_sightings * 0.9, label = "2012 Mayan Apocalypse", color = "darkgreen", angle = 90, vjust = 1.5, size = 3) +
  scale_x_continuous(limits = c(1990, 2013), breaks = seq(1990, 2013, 5)) +
  scale_y_continuous(limits = c(0, max_sightings)) +
  labs(title = 'UFO Sightings (1990–2012)', x = 'Year', y = 'Sightings')

p10
```
:::

By 2012, public fascination with UFOs had reached a cultural high point - reflected in that year’s record number of reported sightings. @cr-ufo-10

::: {#cr-ufo-11}
```{r fig.retina=1, echo=FALSE, warning=FALSE, message=FALSE}
#| fig-width: 6
#| fig-height: 5
#| fig-dpi: 100

old_data <- ufo_by_year %>% filter(year <= 2010)
new_data <- ufo_by_year %>% filter(year >= 2010 & year <= 2012)

label_2012 <- ufo_by_year %>% filter(year == 2012)

p11 <- ggplot() +
  geom_line(data = old_data, aes(x = year, y = sightings), color = "lightblue", linewidth = 1.2) +
  geom_line(data = new_data, aes(x = year, y = sightings), color = "steelblue", linewidth = 1.2) +
  geom_text(
    data = label_2012,
    aes(x = year, y = sightings, label = paste0("Peak: ", sightings)),
    vjust = -1, color = "steelblue", size = 3
  ) +
  scale_x_continuous(limits = c(1990, 2013), breaks = seq(1990, 2013, 5)) +
  scale_y_continuous(limits = c(0, max_sightings)) +
  labs(title = 'UFO Sightings (1990–2012)', x = 'Year', y = 'Sightings')

p11
```
:::

While sightings fluctuated in the years that followed, the buildup to 2012 marked a cultural tipping point, where UFOs moved from fringe curiosity to mainstream conversation. @cr-ufo-12

::: {#cr-ufo-12}
```{r fig.retina=1, echo=FALSE, warning=FALSE, message=FALSE}
#| fig-width: 6
#| fig-height: 5
#| fig-dpi: 100

old_data <- ufo_by_year %>% filter(year <= 2012)
new_data <- ufo_by_year %>% filter(year >= 2012 & year <= 2013)

label_2012 <- ufo_by_year %>% filter(year == 2012)

p12 <- ggplot() +
  geom_line(data = old_data, aes(x = year, y = sightings), color = "lightblue", linewidth = 1.2) +
  geom_line(data = new_data, aes(x = year, y = sightings), color = "steelblue", linewidth = 1.2) +
  scale_x_continuous(limits = c(1990, 2013), breaks = seq(1990, 2013, 5)) +
  scale_y_continuous(limits = c(0, max_sightings)) +
  labs(title = 'UFO Sightings (1990–2013)', x = 'Year', y = 'Sightings')

p12
```
:::
::::::::::::::::

**Conclusion:** UFO sighting reports are not merely records of unexplained phenomena - they reflect broader cultural and societal forces. Our visualizations show that spikes in sightings often align with media attention and conspiracies, suggesting that perception plays a critical role in these encounters. By visualizing when and why these reports occurred, we gain insight not just into the data, but into the evolving relationship between society and belief. Ultimately, the story of UFO sightings is as much about us as it is about what’s in the sky.

```{=html}
<script>
// helper function for executing JavaScript code when a closeread chunk appears
onShow = (selector, callback) => {
  const target = document.querySelector(selector);
  
  const observer = new MutationObserver((mutationsList) => {
    for (const mutation of mutationsList) {
      if (mutation.type === 'attributes' && mutation.attributeName === 'class') {
        if (target.className.includes("cr-active")) {
          callback();
        }
      }
    }
  });
  
  observer.observe(target, { attributes: true, attributeOldValue: true });  
}

// if you want a GIF to restart at the beginning whenever its chunk appears,
// use something like this:
gapminderImg = document.querySelector("#cr-gapminder img");
onShow('#cr-gapminder', () => {
  // this seems silly but it works: by re-assigning the image source to itself,
  // the corresponding animated GIF is restarted at the beginning.
  gapminderImg.src = gapminderImg.src;
})
</script>
```
